Type-Safe Modular Parsing

### 2017-01-10 paper/sections/Inheritance.tex
Polishing section 4 and 5
Revised S3

\section{Modularizing Parsers with Inheritance}\label{sec:inheritance}

This section introduces the problem of semantic modular parsing which motivates our work, and an initial solution using only standard inheritance in OO.

@@

In the rest of Section~\ref{sec:inheritance}, we will introduce an solution for modular parsing with Extension 1 above, using only standard inheritance in OO. Extension 2 will be discussed in Section~\ref{sec:algebrasandparsing}.

\subsection{Inheritance-Based Parser}

Let's firstly consider Extension 1, which is to extend the syntax as well as the parser. As an example, we continue with the expression language of literals and additions in Section~\ref{subsec:packratparsing}. We introduce variables as a new case as below.

@@

It is easy to extend the corresponding AST in an OO way:

@@

Unfortunately, it fails to parse some expressions like \inlinecode{"1 + x"}, which is obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

\subsection{Overriding for Extensibility}

@@

Notice that Scala has a linearized-style multiple inheritance for traits. It can be very helpful when composing several languages. Suppose now
we want to compose the parsers for expressions from pre-defined languages \inlinecode{LanguageA} and \inlinecode{LanguageB} using alternative.
The use of keyword \inlinecode{super} is able to
specify the implementation from inheritance, like:

@@

As we demonstrated, inheritance is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs with Extension 1.

>>>>

\section{Parsing OO ASTs with Multiple Inheritance}\label{sec:inheritance}

Before we address the problem of full modular parsing, we first
address a simpler problem: how to parse Object-Oriented ASTs. To solve
this problem we employ multiple inheritance, which is supported in
Scala via traits.

\subsection{The Mismatch Between OO ASTs and Grammars}
Object-oriented ASTs are ASTs encoded via standard OO techniques, 
such as the {\sc Composite} or {\sc Interpreter} patterns~\cite{gamma94design}. Such ASTs 
fix the operations required in the AST, but allow the easy addition of new 
language constructs. This is in contrast with ASTs modelled using 
functional programming algebraic datatypes or OO style Visitors, where 
adding new operations is easy, but adding new language constructs is hard.
The AST in Figure~\ref{} provides a simple example of an OO AST. The same 
AST could be encoded using algebraic datatypes (or alternatively Visitors) 
as:

\begin{lstlisting} 
data Exp = Lit Int | Add Exp Exp
\end{lstlisting}

Algebraic datatypes and grammars fit well together. Indeed there are
obvious similarities in declaring a grammar, and declaring an
algebraic datatype. It is difficult to add new language
constructs to both grammars and ASTs based on algebraic datatypes. However,
ASTs modelled with algebraic datatypes naturally allow for the easy
addition of operations. So using such ASTs at least preserves one
dimension of extensibility.

Unfortunatelly, when parsing OO style ASTs, there are difficulties with
the two dimensions of extensibility. On the one hand it is naturally hard to 
add new operations to the AST, as usual in the OO style. On the hand 
grammars are not modularly extensible, thus the addition of new
language constructs is problematic. Therefore,
it is perhaps not a suprise that parser generators that also generate 
AST code (auch as ANTLR~\cite{}), normally generate code based on 
visitors. Clearly this seems a better choice since generating code 
based on the {\sc Interpreter} or {\sc Composite} pattern would 
simply be too inflexible, due to the dual limitation on extensibility.

In the remainder of this section we show a technique for 
defining \emph{extensible} parsing code. That is new 
language constructs can be modularly added without changing 
existing parsing code. The technique employs standard OO 
style mechanisms such as subtyping and inheritance, as well as, 
Packrat parsing combinators.

\begin{comment}
This section introduces the problem of semantic modular parsing which motivates our work, and an initial solution using only standard inheritance in OO.

@@

In the rest of Section~\ref{sec:inheritance}, we will introduce an solution for modular parsing with Extension 1 above, using only standard inheritance in OO. Extension 2 will be discussed in Section~\ref{sec:algebrasandparsing}.
\end{comment}

\subsection{Extensible Parsing via Inheritance}

To illustrate how to modularly add new language constructs and the 
corresponding parsing code, we continue with the expression language of literals and additions in Section~\ref{subsec:packratparsing}. We introduce variables as a new case:

@@

It is easy to extend the corresponding OO AST in a modular way:

@@

Unfortunately, this fails to parse some expressions like \inlinecode{"1 + x"}, which are obviously valid in the new grammar.
The reason is that \inlinecode{pAdd} makes two recursive calls to parse sub-expressions, by using \inlinecode{pExpr}, which
covers both cases in the old grammar. But the newly added case \inlinecode{pVar} is not observed by the recursive \inlinecode{pExpr},
hence the parser does not work as expected. It is possible to build the correct parser by replacing the recursive call in \inlinecode{pAdd} with \inlinecode{pVarExpr}.
However, modification on existing code sacrifices separate compilation, as mentioned by the Expression Problem.

\subsection{Overriding for Extensibility}\label{subsec:overriding}

@@

\subsection{Multiple Extensions and Independent Extensibility}
A nice feature of Scala is its support for a form of multiple inheritance.
Scala has a linearized-style multiple inheritance for traits~\cite{}. 
This can be very helpful when composing several languages, which 
may have been independently developed. Suppose now
we want to compose the parsers for expressions from pre-defined
languages \inlinecode{LanguageA} and \inlinecode{LanguageB} using
alternative.\bruno{Don't be abstract! Use concrete examples. You 
already have Var, maybe you can add another language extension for 
boolean literals. Show the code for the boolean literals parsing as
well as the composition code.}
The use of keyword \inlinecode{super} is able to
specify the implementation from inheritance, like:

@@

Interestingly note that, due to the use of multiple inheritance, we 
need two different super calls.\bruno{expand here. People may not 
be familiar with scala super calls, you have to explain what the 
syntax does.}
As we demonstrated, inheritance is the key technique to obtain semantic modularity.
It enables type-safe code reuse and separate compilation for parsing OO style ASTs.

### 2017-01-10 paper/sections/PackratParsing.tex
Revised S3

This section discusses the algorithmic challenges introduced by modular parsing and argues that parser combinators from Packrat parsing
are suitable to address those challenges.

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of them are
problematic in a modular setting.

>>>>

This section discusses the algorithmic challenges introduced by
modular parsing and argues that Packrat parser combinators 
are suitable to address those challenges. The algorithm challenges 
are important because they dictate how modular parsers should be
written. Generally speaking, the methodology we propose is that:

\begin{itemize}

\item {\bf Modular parsers should be left-recursive;}

\item {\bf Modular parsers should use a longest match composition operator.}

\end{itemize}

Moreover, the underlying parsing formalism should make backtracking 
cheap, since \emph{backtracking is pervasive in modular parsing}.

\subsection{Algorithmic Challenges of Modularity}\label{subsec:challenges}
At a first look, parser combinators are very suitable for modular parsing, because of two reasons. Firstly, they are naturally modular. The manner of using them is to write small parsers and use combinators to composed them together. The construction procedure is explicit and fully controlled by the programmer. Secondly, each parser combinator is represented by a piece of code, and also are the parsers it takes. Thus in a statically typed programming language they can be statically type-checked.
Unfortunately many parser combinators have important limitations.
In particular several parser combinators,
including the famous Parsec~\cite{Leijen2001} library, require
programmers to manually do \textit{left-recursion elimination}, \textit{longest match composition}, and
require significant amounts of \textit{backtracking}. All of those are
problematic in a modular setting.
